---
title: "RAV Sex Prediciton RF"
author: "Yoon Ji Jung"
date: "2025-05-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
## Packages -------------
suppressPackageStartupMessages({
  # BiocManager
  library(GenomicSuperSignature)
  library(curatedTCGAData)
  library(MultiAssayExperiment)
  library(TCGAutils)
  library(ComplexHeatmap)
  
  # CRAN
  library(tidyverse) # includes dplyr, ggplot2, magrittr, tidyr
  library(magick)
  library(wordcloud)
  library(ztable)
  library(metafolio)
  library(randomForest)
  library(caret)
  library(pROC)
  library(rstatix)
})
```

1. Import RAVmodel
```{r}
RAVmodel <- getModel('C2', load=TRUE)
```


2. Import sex-specific TCGA datasets and subset training and test datasets
```{r}
tcga <- curatedTCGAData(diseaseCode = c('BRCA', 'UCEC', 'OV', 'PRAD'),
                        assays = 'RNASeq2Gene',
                        version = '2.1.1',
                        dry.run = FALSE)
sampleTables(tcga)
tcga <- TCGAsplitAssays(tcga, c("01", "11"))
br_cancer <- getWithColData(tcga,
                              '01_BRCA_RNASeq2Gene-20160128')
ut_cancer <- getWithColData(tcga,
                              '01_UCEC_RNASeq2Gene-20160128')
ov_cancer <- getWithColData(tcga,
                              '01_OV_RNASeq2Gene-20160128')
pr_cancer <- getWithColData(tcga,
                            '01_PRAD_RNASeq2Gene-20160128')

sex_tcga <- cbind(br_cancer, ut_cancer, ov_cancer, pr_cancer)
assay(sex_tcga) <- log2(assay(sex_tcga) + 1)
table(sex_tcga@colData@listData$patient.gender)

# 80% used for training
set.seed(6)
n_samples <- nrow(sex_tcga)
train_sample_index <- sample(seq_len(n_samples), round(n_samples*0.8))
train_sex_tcga <- sex_tcga[train_sample_index,]

# 20% used for testing
test_sex_tcga <- sex_tcga[-train_sample_index,]
```

3. Find top n validated RAVs for sex-specific TCGA (pearson corr. coeff > 0.45)
```{r}
train_sex_tcga_meta <- colData(train_sex_tcga)
train_sex_tcga_gender <- train_sex_tcga@colData@listData$patient.gender
train_sex_tcga_val <- validate(train_sex_tcga, RAVmodel)
heatmapTable(train_sex_tcga_val, RAVmodel, num.out=15)
top_ravs_sex_tcga <- rownames(train_sex_tcga_val %>% filter(score >= 0.45))
```

4. Calculate sample score and run wilcoxon tests between sample scores and sex
```{r}
sc_sex_tcga_train <- calculateScore(train_sex_tcga, RAVmodel)
sc_sex_tcga_train_subset <- sc_sex_tcga_train[, top_ravs_sex_tcga] %>% as.data.frame()
sc_sex_tcga_train_subset$sex <- as.factor(train_sex_tcga_gender)

wilcoxon_pvalues <- list()
for (i in seq_len(ncol(sc_sex_tcga_train_subset)-1)) {
    RAV <- colnames(sc_sex_tcga_train_subset)[i]
    ## wilcoxon test
    wilcox_test <- wilcox.test(sc_sex_tcga_train_subset[, i] ~ sc_sex_tcga_train_subset$sex, alternative="two.sided")
    wilcoxon_pvalues[[RAV]] <- wilcox_test$p.value
    #print(wilcox_test$p.value)
}

# calculating effect sizes
long_sc_sex_tcga_train_subset <- sc_sex_tcga_train_subset %>%
  pivot_longer(
    cols = -sex,                  # all columns except 'sex'
    names_to = "RAV",
    values_to = "score"
  )

wilcox_effect_sizes <- long_sc_sex_tcga_train_subset %>%
  group_by(RAV)
wilcox_effect_sizes <- wilcox_effect_sizes %>% wilcox_effsize(score ~ sex)
wilcox_effect_sizes_filtered <- wilcox_effect_sizes %>% filter(magnitude != "small")

wilcoxon_pvalues
sex_significant_ravs <- c("RAV27", "RAV58", "RAV99", "RAV152", "RAV220", "RAV221", "RAV338",  "RAV357", "RAV468", "RAV503", "RAV504", "RAV531", "RAV532", "RAV683", "RAV711", "RAV1067", "RAV1076", "RAV1241", "RAV1543", "RAV1575", "RAV2514")

sex_significant_ravs2 <- wilcox_effect_sizes_filtered$RAV
```



5. Import non-sex-specific TCGA datasets
```{r}
tcga_rf <- curatedTCGAData(diseaseCode = c('COAD', 'STAD', 'LUAD'),
                        assays = 'RNASeq2Gene',
                        version = '2.1.1',
                        dry.run = FALSE)
sampleTables(tcga_rf)
tcga_rf <- TCGAsplitAssays(tcga_rf, c("01", "11"))

coad_cancer <- getWithColData(tcga_rf,
                              '01_COAD_RNASeq2Gene-20160128')
stad_cancer <- getWithColData(tcga_rf,
                              '01_STAD_RNASeq2Gene-20160128')
luad_cancer <- getWithColData(tcga_rf,
                              '01_LUAD_RNASeq2Gene-20160128')

comb_non_sex_tcga <- cbind(coad_cancer, stad_cancer, luad_cancer)
assay(comb_non_sex_tcga) <- log2(assay(comb_non_sex_tcga) + 1)
table(comb_non_sex_tcga@colData@listData$patient.gender)

# 80% used for training
set.seed(7)
n_samples2 <- nrow(comb_non_sex_tcga)
train_sample_index2 <- sample(seq_len(n_samples2), round(n_samples2*0.8))
train_non_sex_tcga <- comb_non_sex_tcga[train_sample_index2,]

# 20% used for testing
test_non_sex_tcga <- comb_non_sex_tcga[-train_sample_index2,]
```

6. Find top n validated RAVs for non-sex-specific TCGA
```{r}
train_non_sex_tcga_meta <- colData(train_non_sex_tcga)
train_non_sex_tcga_gender <- train_non_sex_tcga@colData@listData$patient.gender
train_non_sex_tcga_val <- validate(train_non_sex_tcga, RAVmodel)
heatmapTable(train_non_sex_tcga_val, RAVmodel, num.out=15)
top_ravs_non_sex_tcga <- rownames(train_non_sex_tcga_val %>% filter(score >= 0.45))
```
7. Calculate sample scores and run wilcoxon tests between sample scores and sex
```{r}
sc_non_sex_tcga_train <- calculateScore(train_non_sex_tcga, RAVmodel)
sc_non_sex_tcga_train_subset <- sc_non_sex_tcga_train[, top_ravs_non_sex_tcga] %>% as.data.frame()
sc_non_sex_tcga_train_subset$sex <- as.factor(train_non_sex_tcga_gender)

wilcoxon_pvalues2 <- list()
for (i in seq_len(ncol(sc_non_sex_tcga_train_subset)-1)) {
    RAV <- colnames(sc_non_sex_tcga_train_subset)[i]
    ## wilcoxon test
    wilcox_test <- wilcox.test(sc_non_sex_tcga_train_subset[, i] ~ sc_non_sex_tcga_train_subset$sex, alternative="two.sided")
    wilcoxon_pvalues2[[RAV]] <- wilcox_test$p.value
}

# calculating effect sizes
long_sc_non_sex_tcga_train_subset <- sc_non_sex_tcga_train_subset %>%
  pivot_longer(
    cols = -sex,                  # all columns except 'sex'
    names_to = "RAV",
    values_to = "score"
  )

wilcox_effect_sizes_non_sex <- long_sc_non_sex_tcga_train_subset %>%
  group_by(RAV)
wilcox_effect_sizes_non_sex <- wilcox_effect_sizes_non_sex %>% wilcox_effsize(score ~ sex)
wilcox_effect_sizes_non_sex_filtered <- wilcox_effect_sizes_non_sex %>% filter(magnitude != "small")

wilcoxon_pvalues2
non_sex_tcga_sex_significant_ravs <- c("RAV64", "RAV188", "RAV220", "RAV324", "RAV438", "RAV504" , "RAV758", "RAV831", "RAV868", "RAV1019", "RAV1319", "RAV2489", "RAV2528",
"RAV2528")

non_sex_tcga_sex_significant_ravs2 <- wilcox_effect_sizes_non_sex_filtered$RAV

# No RAVs with moderate to high effect sizes for sex from the non-sex specific training dataset

```

8. Find overlapping sex-specific and non-sex-specific TCGA sex-significant RAVs
6/6/25 edit: decided to proceed with only RAVs with mod-high effect sizes from wilcoxon tests 
conducted on sex-specific tcga training set only
```{r}
#overlap_sex_ravs <- intersect(non_sex_tcga_sex_significant_ravs, sex_significant_ravs)
```

# Sex Prediction RF Model Building and Testing
1. Subset Sample Scores from sex-specific TCGA data for RAV220 and RAV504
```{r}
sc_sex_tcga_train_subset2 <- sc_sex_tcga_train_subset[, sex_significant_ravs2]
sc_sex_tcga_train_subset2$sex <- as.factor(train_sex_tcga_gender)
```

2. Build RF Model
```{r}
predict_sex_rfmodel <- randomForest(sex ~ RAV1076 + RAV1241 + RAV152 + RAV220 + RAV221 + RAV2514, sc_sex_tcga_train_subset2, proximity = TRUE)
```

3. Calculate samples scores of test data from sex-specific tcga
```{r}
sc_sex_tcga_test_gender <- test_sex_tcga@colData@listData$patient.gender
sc_sex_tcga_test <- calculateScore(test_sex_tcga, RAVmodel)
sc_sex_tcga_test_subset <- sc_sex_tcga_test[,sex_significant_ravs2] %>% as.data.frame()
sc_sex_tcga_test_subset$sex <- as.factor(sc_sex_tcga_test_gender)
```

4. Test RF model using test data from sex-specific tcga
```{r}
test_RF <- predict(predict_sex_rfmodel, sc_sex_tcga_test_subset, type = "prob")

##ROC curves
roc_sex_rf <- roc(sc_sex_tcga_test_subset$sex, test_RF[, "female"])
## Plot ROC curves
plot(roc_sex_rf, col = "blue")

##feature importance
importance(predict_sex_rfmodel)
varImpPlot(predict_sex_rfmodel)

# confusion matrix
confusionMatrix(predict(predict_sex_rfmodel, sc_sex_tcga_test_subset), sc_sex_tcga_test_subset$sex)
```

# Test data using non-sex specific TCGA data
1. Calculate samples scores of test data from non-sex-specific tcga
```{r}
sc_non_sex_tcga_test_gender <- test_non_sex_tcga@colData@listData$patient.gender
sc_non_sex_tcga_test <- calculateScore(test_non_sex_tcga, RAVmodel)
sc_non_sex_tcga_test_subset <- sc_non_sex_tcga_test[,sex_significant_ravs2] %>% as.data.frame()
sc_non_sex_tcga_test_subset$sex <- as.factor(sc_non_sex_tcga_test_gender)
```

2. Test RF model using non-sex specific TCGA data
```{r}
test_RF2 <- predict(predict_sex_rfmodel, sc_non_sex_tcga_test_subset, type = "prob")
# confusion matrix
confusionMatrix(predict(predict_sex_rfmodel, sc_non_sex_tcga_test_subset), sc_non_sex_tcga_test_subset$sex)
```

