---
title: "MSI Status Prediction RF Model"
author:
  - Sehyun Oh
date: "`r format(Sys.time(), '%B %d, %Y')`"
format:
    html:
        fontsize: 14pxs
        toc: true
        top-depth: 3
abstract: "Using the TCGA-COAD cancer samples for training"
---

```{r load_packages, echo=FALSE}
suppressPackageStartupMessages({
    library(caret)      # For model training and evaluation
    library(pROC)       # For ROC curves
    library(ROSE)       # For handling imbalanced data
    library(glmnet)     # For regularized logistic regression
    library(randomForest)
    library(xgboost)
    library(GenomicSuperSignature)
    library(tidyverse)
})

source("~/Projects/InferPhenoCuration/R/calculate_auc.R")
source("~/Projects/InferPhenoCuration/R/plot_roc.R")
```

```{r}
RAVmodel <- getModel('C2', load=TRUE)
```

# 1. Data Preprocessing

```{r}
## Samples with MSI status info
combined_data <- readRDS("data/combined_data.rds") # combined cancer samples from three TCGA datasets
target_attr <- "patient.microsatellite_instability_test_results.microsatellite_instability_test_result.mononucleotide_and_dinucleotide_marker_panel_analysis_status"
combined_data_msi <- combined_data[,combined_data[[target_attr]] %in% c("msi-h", "mss")]

## Predictor RAVs
RAVs_combinedTCGA <- c(
    517, 220, 2109, 1303, 324, 438, 868, # RAVs that have statistically significant pairwise wilcoxon p-values of mss vs msi-h
    834, 190, 1166, # RAVs with significant KW test statistic (p-value < 0.05) for COAD
    2344, # significant KW test value for STAD, includes 324, 868, 517 above
    357) # UCEC KW test value (p-value = 0.056)
RAVmodel_sub <- RAVmodel[,RAVs_combinedTCGA]

## Calculate sample scores
sampleScore <- calculateScore(assay(combined_data_msi), RAVmodel_sub)
```

The `data` object with the predictors (i.e., sample scores from 12 RAVs) and 
response (i.e., MSI status) variable.

```{r}
data <- as.data.frame(sampleScore)
data$status <- colData(combined_data_msi)[[target_attr]]
data$status <- ifelse(data$status == "msi-h", "MSI", "MSS")
data$status <- factor(data$status, levels = c("MSS", "MSI")) # Convert outcome to factor
```

## Split data into training and testing sets
```{r}
table(colData(combined_data)$admin.disease_code)
disease_code <- colData(combined_data)[c("patientID", "admin.disease_code")]

## Train dataset: TCGA-COAD
coad_id <- disease_code %>% as.data.frame %>% dplyr::filter(admin.disease_code == "coad") %>% pull(patientID)
train_coad <- data[rownames(data) %in% coad_id,]

## Test datasets
stad_id <- disease_code %>% as.data.frame %>% dplyr::filter(admin.disease_code == "stad") %>% pull(patientID)
test_stad <- data[rownames(data) %in% stad_id,]
ucec_id <- disease_code %>% as.data.frame %>% dplyr::filter(admin.disease_code == "ucec") %>% pull(patientID)
test_ucec <- data[rownames(data) %in% ucec_id,]
```



# 2. Handle class imbalance using `ROSE`
`ROSE` is used for handling class imbalance (alternatives include `SMOTE` from the *DMwR* package)

```{r}
# balanced_train <- ROSE(status ~ ., data = train_coad)$data
```


# 3. Basic Logistic Regression
Both logistic regression and random forest are implemented for comparison

```{r}
# ## With cross-validation
# ctrl <- trainControl(method = "cv", 
#                      number = 5,
#                      classProbs = TRUE,
#                      summaryFunction = twoClassSummary)
# 
# ## Train logistic regression
# log_model <- train(status ~ .,
#                    data = balanced_train,
#                    method = "glm",
#                    family = "binomial",
#                    trControl = ctrl,
#                    metric = "ROC")
```


# 4. Random Forest with class weights
Feature importance is assessed through random forest

```{r}
rf_model <- randomForest(status ~ .,
                         data = train_coad,
                         ntree = 500,
                         classwt = c("MSS" = 1, "MSI" = 3))
```


# 5. Model Evaluation
Performance evaluation includes ROC curves and confusion matrices


## TCGA-STAD test data
```{r}
dat <- test_stad
rf_pred <- predict(rf_model, dat, type = "prob")
plot_roc_pROC(dat$status, rf_pred[, "MSI"], 
              paste0("TCGA-STAD MSI Status ROC (n = ", nrow(dat),")")) # update the plot title

roc_stad <- roc(dat$status, rf_pred[, "MSI"])
```

## TCGA-UCEC test data
```{r}
dat <- test_ucec
rf_pred <- predict(rf_model, dat, type = "prob")
plot_roc_pROC(dat$status, rf_pred[, "MSI"], 
              paste0("TCGA-UCEC MSI Status ROC (n = ", nrow(dat),")")) # update the plot title

roc_ucec <- roc(dat$status, rf_pred[, "MSI"])
```

```{r fig.width=5, fig.height=5}
## Combined ROC curves
# plot(roc_rf, col = "blue")
# lines(roc_coad, col = "red")
plot(roc_stad, col = "orange")
lines(roc_ucec, col = "yellow")

legend("bottomright", legend = c("TCGA-STAD", "TCGA-UCEC"),
       col = c("orange", "yellow"), lwd = 2)
```




# 6. Feature Importance (for Random Forest)
```{r}
importance(rf_model)
varImpPlot(rf_model)
```


# 7. Confusion Matrix
```{r collapse=FALSE}
confusionMatrix(predict(rf_model, test_stad), test_stad$status) # TCGA-STAD (n = 99)
confusionMatrix(predict(rf_model, test_ucec), test_ucec$status) # TCGA-UCEC (n = 44)
```

